Population structure and phenotypic variation of *Sclerotinia sclerotiorum* from dry bean in the United States

### 2017-08-01 doc/manuscript/manuscript.Rmd
rearrange manuscript
 - remove "phenotypic"
 - make AMOVA table auto-generating
 - Update aggressiveness assessment (#8)
 - narrow widths for half-page plots (to account for center column)
 - add aggressiveness figure
 - change aggressiveness to be region-focused
 - move MCG distribution figure to supplemental
 - Move MLG-MCG correlation to Variable Assessment
 - Add statements about cultivar DAPC results
 - begin discussion onf population differentiation

## Assessing importance of categorical and phenotypic variables  {-}

### Distance-based redundancy analysis {-}

@@

### Aggressiveness assessment {-}

We used analysis of variance (ANOVA) to assess if MCGs differed in aggressiveness.
To minimize complications due to small sample sizes, we chose the top 10 MCGs, representing 57% of the isolates collected. Between-MCG differences were assessed using Tukey's HSD test with an alpha of 0.05 using the `HSD.test()` function in the package *agricolae* version 1.2.4 [@mendiburu2015agricolae].

@@

A total of 366 isolates were collected from 2003 to 2012 (except 2006 and 2011) from diseased dry bean plants in 11 states in the US, Australia, Mexico, and France (Table \@ref(tab:isolate-table)).
Over 50% of the isolates came from four states, MI (62), ND (60), WA (59), NE (47). 
Four regions had fewer than 10 isolates, Australia (6), WI (2), NY (1), ID (1).
After clone-correction on the hierarchy of Region/Source/Host/Year, we were left with 318 isolates representing 165 MLGs (Table \@ref(tab:iatab)).
Analysis of the genotype accumulation curve showed no clear plateau, indicating that we would likely obtain more multilocus genotypes if we were to collect more loci. The number of observed alleles per locus ranged from 2 to 10 with an average of 6.27 (Table \@ref(tab:locus-stats)).

@@

## Phenotypic Variable Assessment {-}

@@

### Mycelial Compatibility Groups {-}

We observed 87 mycelial compatibility groups (MCG), the most abundant of which (MCG 5) represented 73 isolates over 37 multilocus genotypes (MLG) (Fig. \@ref(barplot)). 
Our assessment of aggressiveness association with MCGs showed a significant effect (P < 0.001) indicating that MCG 44 is, on average, more aggressive than other MCGs (Table \@ref(tab:mcg-aggressiveness)).

@@

![
Biplot showing five most influential explanatory variables (arrows) overlayed on the first two eigenvectors of distance based redundancy analysis of *S. sclerotiorum* isolates. 
The length of the arrows are directly proportional to the strength of the correlation between explanatory and molecular variables. 
Points represent the 318 clone-corrected samples in ordination space. 
](../../results/figures/publication/Figure7Z.pdf){#rda-biplot width=50%}

@@

@@

@@

We observed a total of 165 multilocus genotypes (MLG) (215 with 16 loci).
The most abundant MLG represented 27 isolates (Table \@ref(tab:mlg-table)) across five regions (NE, MI, WA, CO, and ND). 
The genotypes were relatively evenly distributed within regions with moderate to high diversity (Table \@ref(tab:iatab)). 
Of the 165 MLGs, 76 (46%) were found across all geographic regions except WI (2), ID (1), and Mexico (18) (Fig. \@ref(community-graph)).
Community analysis of the MLGs shared between populations revealed 4 communities with a modularity of 0.17 (Fig. \@ref(community-graph)): A coastal community (CA, OR, WA, and NY), a mid-west community (CO, NE, ND, MI), and an international community (MN, France, Australia).
Analysis with 16 loci resulted in the removal of the NY node because it no longer shared a genotype with OR, but the same overall community structure was present with a modularity of 0.2 (Fig. \@ref(community-graph16)).
The international community appears to be driven by MLG 4, which is shared between all three populations and has a $P_{sex}$ value of 2.87 x 10^-5^, in contrast to the abundant MLG 25, which has a $P_{sex}$ value of 0.0168.

@@

The analysis of molecular variance (AMOVA) over the hierarchy of Source and Region showed no significant variation between white mold nurseries and non-white mold nurseries after accounting for variation between regions (Table \@ref(tab:amova)).
Because Mexico shared no genotypes with any of the other populations and we suspected that it was contributing to some of the observed variance, we re-ran AMOVA without the Mexican isolates and found similar decreased, but still significant levels of differentiation between regions.
In contrast, when we compared the three cultivars, Beryl, Bunsi, and G122, we found no evidence of population differentiation (Supplementary Data).

-------------------------------------------------------------------------------
Hierarchy               d.f.        S.S.       % variation    $\Phi$ statistic  
-------------------- ---------- ------------- -------------- ------------------
Between Source         1 (1)    0.897 (0.753) -0.158 (0.116)   0.136 (0.0965)

Between Region        20 (19)     12 (8.91)    13.8 (9.53)     **0.138 (0.0955)**   
within Source

Within Region        296 (282)   56.7 (55.1)   86.4 (90.3)   **-1.58e^-3^ (1.16e^-3^)**

-------------------------------------------------------------------------------

Table: (\#tab:amova) Analysis of Molecular Variance on Bruvo's genetic distance showing little differentiation between White Mold Nurseries and other sources. The hierarchy was constructed as Source/Region where source is defined as belonging to a white mold nursery or not. Bold $\Phi$ values indicate significant values. Numbers in parentheses represent AMOVA run without Mexican isolates included. S.S. = Sum of Squares, d.f. = degrees of freedom.

@@

Discriminant Analysis of Principal Components (DAPC) was performed with the first 21 principal components, representing 88% of the total variance. 
The first discriminant axis (representing 64% of the discriminatory power) separated the centroid for the Mexico isolates from the rest of the data, indicating strong differentiation (Fig. \@ref(DAPC)b). 
The second discriminant axis separated the centroid for the Californian isolates.
The mean population assignment probabilities for all populations with n > 10 showed that only isolates from Mexico, CA, and France had > 50% probabilities of being reassigned to their source populations (Fig. \@ref(DAPC)a).

@@

## Phenotypic Variable Importance {-}

@@

Of interest in this study was testing the association between multilocus genotypes (MLGs) and mycelial compatibility groups (MCGs), as the latter have historically been used as a proxy for clonal lineages [@leslie1993fungal; @otto-hanson2011variation; @kohn1990mycelial; @schafer2006optimized; @kohn1995clonal; @carbone1999patterns]. 
Our results do not support this assumption.
It can be seen in Fig. \@ref(barplot) that the most abundant MCG contains several MLGs, but the diversity of those MLGs are low as indicated by the evenness (transparency), which indicates that there is one dominant MLG (MLG 25).
What is not shown in Fig. \@ref(barplot) is the MLGs that are shared between MCGs. 
This is illustrated in both table \@ref(tab:mlg-table) and Fig. \@ref(biggraph). 
It could be argued, however that MLG 25, with its high value of $P_{sex}$ represents different true MLGs across the five MCGs it occupies, but this doesn't account for the overall structure of Fig. \@ref(biggraph) where, for example, MLG 75 ($P_{sex} = 1.81e^-4$) is compatible with 58 other genotypes through three MCG when the population structure of *S. sclerotiorum* is known to be clonal.

@@

@@

```{r table2, echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
suppressPackageStartupMessages(library("tidyverse"))
read_csv("../../results/tables/MCG-aggressiveness.csv", 
         na = "NA", # spaces read in as spaces
         col_types = cols(.default = col_character())) %>%
  dplyr::rename(MCG = trt) %>%
  dplyr::rename(Group = M) %>%
  dplyr::rename(`Mean Aggressiveness` = means) %>%
  huxtable::as_huxtable(add_colnames = TRUE) %>%
  huxtable::set_number_format(row = huxtable::everywhere, col = 1, 0) %>% 
  huxtable::set_col_width(c(0.05, 0.3, 0.08)) %>%
  huxtable::set_align(row = huxtable::everywhere, col = 2, "right") %>%
  huxtable::print_md(max_width = 31)
```

Table: (\#tab:mcg-aggressiveness) Mean aggressiveness ratings for the ten most abundant mycelial compatibility groups (MCG) with groupings according to 95% family-wise confidence interval.

>>>>

## Assessing Importance of Variables  {-}

### Distance-based Redundancy Analysis {-}

@@

### Aggressiveness Assessment {-}

We used analysis of variance (ANOVA) to assess if aggressiveness could be explained by Region, mycelial compatibility group, or multilocus genotype. 
To minimize complications due to small sample sizes, we chose the top 10 MCGs, representing 56.5% of the isolates collected, the top 10 MLGs representing 26.7% of the isolates, and populations with more than five isolates.
Between-group differences were assessed using Tukey's HSD test with an alpha of 0.05 using the `HSD.test()` function in the package *agricolae* version 1.2.4 [@mendiburu2015agricolae].

@@

A total of 366 isolates were collected from 2003 to 2012 (except 2006 and 2011) from diseased dry bean plants in 11 states in the US, Australia, Mexico, and France (Table \@ref(tab:isolate-table)).
Over 50% of the isolates came from four states, MI (62), ND (60), WA (59), NE (47). 
Four regions had fewer than 10 isolates, Australia (6), WI (2), NY (1), ID (1).
We observed 87 mycelial compatibility groups (MCG), the most abundant of which (MCG 5) represented 73 isolates over 37 multilocus genotypes (MLG) (Fig \@ref(barplot)).
After clone-correction on the hierarchy of Region/Source/Host/Year, we were left with 318 isolates representing 165 MLGs (Table \@ref(tab:iatab)).
Analysis of the genotype accumulation curve showed no clear plateau, indicating that we would likely obtain more multilocus genotypes if we were to collect more loci. The number of observed alleles per locus ranged from 2 to 10 with an average of 6.27 (Table \@ref(tab:locus-stats)).

@@

## Variable Assessment {-}

@@

@@

![
Biplot showing five most influential explanatory variables (arrows) overlayed on the first two eigenvectors of distance based redundancy analysis of *S. sclerotiorum* isolates. 
The length of the arrows are directly proportional to the strength of the correlation between explanatory and molecular variables. 
Points represent the 318 clone-corrected samples in ordination space. 
](../../results/figures/publication/Figure7Z.pdf){#rda-biplot width=45%}

@@

### Aggressiveness {-}

<!-- aggressiveness figure -->

![
Boxplot representing aggressiveness (vertical axis) by population (horizontal axis) arranged in descending order of mean aggressiveness for all populations with N > 5. Box boundaries represent 25% and 75% percentiles and whiskers extend to at most 1.5 $\times$ the interquartile range. Horizontal line represents median value. Points represent individual observations whose colors indicate if they came from white mold screening nurseries (wmn) or not (other).
](../../results/figures/publication/aggressiveness.pdf){#aggressiveness-boxplot width=45%}

@@

Our assessment of aggressiveness in association with Region showed a significant effect (P < 1.00e^-4^), with means that ranged from 5.8 (MN) to 4.0 (CA) (Fig. \@ref(aggressiveness-boxplot), Table \@ref(tab:region-aggressiveness)). 
MCGs also showed a significant effect (P < 1.00e^-4^), with means that ranged from 6.0 (MCG 44) to 4.6 (MCG 49).
We additionally found a significant effect for MLGs (P = 7.44e^-4^) for MLGs, with means that ranged from 6.0 (MLG 78) to 4.3 (MLG 140).

### Correlation of Mulitlocus Genotypes and Mycelial Compatibility Groups {-}

@@

We observed a total of 165 multilocus genotypes (MLG) (215 with 16 loci).
The most abundant MLG represented 27 isolates (Table \@ref(tab:mlg-table)) across five regions (NE, MI, WA, CO, and ND). 
The genotypes were relatively evenly distributed within regions with moderate to high diversity (Table \@ref(tab:iatab)). 
Of the 165 MLGs, 76 (46%) were found across all geographic regions except WI (2), ID (1), and Mexico (18) (Fig. \@ref(community-graph)).
Community analysis of the MLGs shared between populations revealed 4 communities with a modularity of 0.17 (Fig. \@ref(community-graph)): A coastal community (CA, OR, WA, and NY), a mid-west community (CO, NE, ND, MI), and an international community (MN, France, Australia).
Analysis with 16 loci resulted in the removal of the NY node because it no longer shared a genotype with OR, but the same overall community structure was present with a modularity of 0.2 (Fig. \@ref(community-graph16)).
The international community appears to be driven by MLG 4, which is shared between all three populations and has a $P_{sex}$ value of 2.87e^-5^, in contrast to the abundant MLG 25, which has a $P_{sex}$ value of 0.0168.

@@

The analysis of molecular variance (AMOVA) over the hierarchy of Source and Region showed no significant variation between white mold nurseries and non-white mold nurseries after accounting for variation between regions (Table \@ref(tab:amova)).
Because Mexico shared no genotypes with any of the other populations and we suspected that it was contributing to some of the observed variance, we re-ran AMOVA without the Mexican isolates and found similar decreased, but still significant levels of differentiation between regions.
In contrast, when we compared the three cultivars, Beryl, Bunsi, and G122, we found no significant differentiation (Supplementary Data).

```{r table4, echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
suppressPackageStartupMessages(library("tidyverse"))
read_csv("../../results/tables/AMOVA-region.csv", 
         na = "NA", # spaces read in as spaces
         col_types = cols(.default = col_character()))  %>%
  rename(ps = `Phi statistic`) %>%
  mutate(ps = gsub("0\\.00(\\d{1})(\\d{2})", "\\1.\\2e^-3^", ps)) %>%
  mutate(ps = case_when(Hierarchy == "Between Source" ~ ps, TRUE ~ paste0("**", ps, "**"))) %>%
  rename(`$\\Phi statistic$` = ps) %>%
  rename(`% variation` = `Percent variation`) %>%
  rename(S.S. = `Sum of Squares`) %>%
  select(-P) %>%
  huxtable::as_huxtable(add_colnames = TRUE) %>% 
  huxtable::set_col_width(c(1.1, 0.6, 0.9, 0.9, 1.4)) %>% 
  huxtable::set_align(huxtable::everywhere, 2:5, "center") %>% 
  huxtable::print_md(max_width = 100)
```

Table: (\#tab:amova) Analysis of Molecular Variance on Bruvo's genetic distance showing little differentiation between White Mold Nurseries and other sources. The hierarchy was constructed as Source/Region where source is defined as belonging to a white mold nursery or not. Bold $\Phi$ values indicate significant values (P < 1.00e^-4^). Numbers in parentheses represent AMOVA run without Mexican isolates included. S.S. = Sum of Squares, d.f. = degrees of freedom.

@@

Discriminant Analysis of Principal Components (DAPC) was performed grouping by Region with the first 21 principal components, representing 88% of the total variance. 
The first discriminant axis (representing 64% of the discriminatory power) separated the centroid for the Mexico isolates from the rest of the data, indicating strong differentiation (Fig. \@ref(DAPC)b). 
The second discriminant axis separated the centroid for the Californian isolates.
The mean population assignment probabilities for all populations with n > 10 showed that only isolates from Mexico, CA, and France had > 50% probabilities of being reassigned to their source populations (Fig. \@ref(DAPC)a). 
DAPC grouping by cultivar used the first 20 principal components, representing 89% of the total variance.
The first two discriminant axes failed to separate any of the cultivars where the mean posterior assignment probabilities were 33.97% (G122), 35.95% (Beryl), and 30.08% (Bunsi).

@@

## Variable Importance {-}

@@

Of interest in this study was testing the association between multilocus genotypes (MLGs) and mycelial compatibility groups (MCGs), as the latter have historically been used as a proxy for clonal lineages [@leslie1993fungal; @otto-hanson2011variation; @kohn1990mycelial; @schafer2006optimized; @kohn1995clonal; @carbone1999patterns]. 
Our results do not support this assumption.
It can be seen in Fig. \@ref(barplot) that the most abundant MCG contains several MLGs, but the diversity of those MLGs are low as indicated by the evenness (transparency), which indicates that there is one dominant MLG (MLG 25).
What is not shown in Fig. \@ref(barplot) is the MLGs that are shared between MCGs. 
This is illustrated in both table \@ref(tab:mlg-table) and Fig. \@ref(biggraph). 
It could be argued, however that MLG 25, with its high value of $P_{sex}$ represents different true MLGs across the five MCGs it occupies, but this doesn't account for the overall structure of Fig. \@ref(biggraph) where, for example, MLG 75 ($P_{sex}$ = 1.81e^-4^) is compatible with 58 other genotypes through three MCG when the population structure of *S. sclerotiorum* is known to be clonal.

@@

## Regional Differentiation {-}

The primary interest of this study was to assess if isolates sampled from white mold screening nurseries represent isolates from producer fields within the region [@steadman2003identification; @otto-hanson2011variation].
According to our AMOVA results, we have evidence for differentiation at the region level, but little to no differentiation within region (Table \@ref(tab:amova)).

@@

```{r table2, echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
suppressPackageStartupMessages(library("tidyverse"))
read_csv("../../results/tables/Region-aggressiveness.csv", 
         na = "NA", # spaces read in as spaces
         col_types = cols(.default = col_character())) %>%
  huxtable::as_huxtable(add_colnames = TRUE) %>%
  huxtable::set_number_format(row = huxtable::everywhere, col = 1, 0) %>% 
  huxtable::set_col_width(c(0.13, 0.25, 0.08)) %>%
  huxtable::set_align(row = huxtable::everywhere, col = 2, "right") %>%
  huxtable::print_md(max_width = 40)
```

Table: (\#tab:region-aggressiveness) Mean aggressiveness ratings for the Regions with greater than 5 samples; groupings according to 95% family-wise confidence interval.
